---
title: "Initial Analysis of Wordbank Data"
author: "George, Alvin, Julien, ..."
format: html
editor: source
---

## Setup
```{r}
knitr::opts_chunk$set(message = FALSE)
```

```{r}
library(tidyverse)
library(wordbankr)
library(glue)
library(gamlss)

theme_set(theme_classic())
```

## Get data 
Get all datasets explicitly labelled "bilingual"
```{r}
bilingual_datasets <- get_datasets() |> 
  filter(str_detect(dataset_origin_name, "Bilingual"))
```

Wrangle exposure data
```{r}
bilingual_data <- get_administration_data(include_demographic_info = TRUE,
                                          include_language_exposure = TRUE) |> 
  filter(dataset_origin_name %in% bilingual_datasets$dataset_origin_name) |>
  unnest(language_exposures, names_sep = "_") |>
  filter(!is.na(language_exposures_language),
         !is.na(language_exposures_exposure_proportion)) |>
  rename(exposure_language = language_exposures_language,
         exposure_proportion = language_exposures_exposure_proportion,
         age_first_exposed = language_exposures_age_of_first_exposure) |>
  select(-c(age_first_exposed, is_norming, date_of_test)) |> 
  filter(str_detect(language, glue("^{exposure_language}")))

# The Armon-Lotem data only have 4 values for exposure_proportion
bilingual_data_clean <- bilingual_data |> 
  filter(dataset_origin_name != "Armon-Lotem_Hebrew_English_Bilingual")
```

## Preprocess data
```{r}
all_instruments <- bilingual_data_clean |> 
  distinct(language, form)
```

Find number of items on each form
```{r}
items <- map2(all_instruments$language, all_instruments$form, get_item_data) |> 
  list_rbind() |> 
  group_by(language, form) |> 
  filter(item_kind == "word") |> 
  summarise(n = n(), .groups = "drop")

bilingual_data_prop <- bilingual_data_clean |> 
  left_join(items, by = join_by(language, form)) |> 
  mutate(prop_prod = production / n,
         prop_prod = case_when(
           prop_prod == 0 ~ .001,
           prop_prod == 1 ~ .999,
           .default = prop_prod
         ),
         child_id = as_factor(child_id))
```

Filter down to just Eng (Am) data for now
```{r}
bilingual_data_prop_en <- bilingual_data_prop |> 
  filter(language == "English (American)")
```

## Run models
Fit GAMLSS model with monotonic spline for exposure
```{r}
gam_nonlinear <- gamlss(prop_prod ~ pbm(age, lambda = 10000) * 
                          pbm(exposure_proportion, lambda = 10000) +
                          re(random = ~ 1 | child_id, level = 0),
                        sigma.formula = ~ pbm(age, lambda = 10000) * 
                          pbm(exposure_proportion, lambda = 10000),
                        data = bilingual_data_prop_en |> 
                          select(prop_prod, age, exposure_proportion, child_id),
                        family = BE,
                        control = gamlss.control(n.cyc = 100))
```

Plot model predictions 
```{r}
pred_params <- expand_grid(age = 17:36, 
                           exposure_proportion = seq(0, 100, length.out = 20),
                           child_id = factor(0))
preds <- predict(gam_nonlinear,
                 newdata = pred_params,
                 type = "response")
gam_nonlinear_preds <- pred_params |> cbind(preds)

ggplot(gam_nonlinear_preds,
       aes(x = age, y = preds, 
           col = exposure_proportion, 
           group = exposure_proportion)) +
  geom_line() +
  labs(x = "Age", y = "Proportion produced", col = "Exposure proportion")
```

```{r}
ggplot(gam_nonlinear_preds,
       aes(x = exposure_proportion, y = preds, 
           col = age, 
           group = age)) +
  geom_line() +
  labs(x = "Exposure proportion", y = "Proportion produced", col = "Age")
```


Compare with linear exposure term
```{r}
gam_linear <- gamlss(prop_prod ~ pbm(age, lambda = 10000) * 
                       exposure_proportion +
                       re(random = ~ 1 | child_id, level = 0),
                     sigma.formula = ~ pbm(age, lambda = 10000) * 
                       exposure_proportion,
                     data = bilingual_data_prop_en |> 
                       select(prop_prod, age, exposure_proportion, child_id),
                     family = BE,
                     control = gamlss.control(n.cyc = 100))

# LRfunc credit: https://sakai.unc.edu/access/content/group/3d1eb92e-7848-4f55-90c3-7c72a54e7e43/public/docs/lectures/lecture18.htm
LRfunc <- function(x, y) {
  LR <- 2 * (logLik(y) - logLik(x))
  df <- attr(logLik(y), "df") - attr(logLik(x), "df")
  p <- 1 - pchisq(LR, df)[1]
  out <- data.frame(LR = LR, df = df, p = p)
  print(out, row.names = F)
}

LRfunc(gam_linear, gam_nonlinear)
```
The model with a nonlinear exposure term is significantly better than the model with a linear exposure term (using a likelihood ratio test). (This result also holds if you use AIC for model selection, but there is no significant difference using BIC.)

Now fitting on data from all current languages
```{r}
gam_nonlinear_all <- gamlss(prop_prod ~ pbm(age, lambda = 10000) * 
                              pbm(exposure_proportion, lambda = 10000) +
                              re(random = ~ 1 | child_id, level = 0) +
                              re(random = ~ 1 | language, level = 0),
                            sigma.formula = ~ pbm(age, lambda = 10000) * 
                              pbm(exposure_proportion, lambda = 10000),
                            data = bilingual_data_prop |> 
                              select(prop_prod, age, exposure_proportion, 
                                     child_id, language),
                            family = BE,
                            control = gamlss.control(n.cyc = 100))
gam_linear_all <- gamlss(prop_prod ~ pbm(age, lambda = 10000) * 
                           exposure_proportion +
                           re(random = ~ 1 | child_id, level = 0) +
                           re(random = ~ 1 | language, level = 0),
                         sigma.formula = ~ pbm(age, lambda = 10000) * 
                           exposure_proportion,
                         data = bilingual_data_prop |> 
                           select(prop_prod, age, exposure_proportion, 
                                  child_id, language),
                         family = BE,
                         control = gamlss.control(n.cyc = 100))
LRfunc(gam_linear_all, gam_nonlinear_all)
```
Not sure why the df difference is negative here (which results in a NaN p-value). Nonetheless AIC and BIC both prefer the nonlinear model.
