---
title: "O1 Covariates (sex, SES) — exploratory"
format: html
params:
  stage2_variant: "linear"
  m_imputations: 10
  run_covariate_models: true
---

```{r, include=FALSE}
library(glue)
library(here)
library(brms)
library(cmdstanr)
library(mice)
library(patchwork)
library(tidyverse)

theme_set(
  theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    )
)

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.width = 8,
  fig.height = 5
)
```

# Stage-2 data with demographics

```{r}
child_data <- readRDS(here("data", "bilingual_hirt_stage2_data.rds")) |>
  filter(!is.na(theta_est))

df_demogs <- readRDS(here("data", "demographics.rds")) |>
  mutate(data_id = as.character(data_id))

demo_cov <- df_demogs |>
  dplyr::select(data_id, sex, caregiver_education)

child_o1 <- child_data |>
  left_join(demo_cov, by = "data_id")

child_o1_clean <- child_o1 |>
  filter(age >= 7) |>
  mutate(
    exposure_proportion = exposure_proportion / 100,
    exp_c = scale(exposure_proportion, scale = FALSE)[, 1],
    age_c = scale(age, scale = FALSE)[, 1]
  )
```

# SES and sex coding

```{r}
edu_levels <- c(
  "None",
  "Primary",
  "Some Secondary",
  "Secondary",
  "Some College",
  "College",
  "Some Graduate",
  "Graduate"
)

child_o1_clean <- child_o1_clean |>
  mutate(
    edu_ord = factor(caregiver_education, levels = edu_levels, ordered = TRUE),
    sex_std = case_when(
      sex %in% c("F", "Female") ~ "Female",
      sex %in% c("M", "Male") ~ "Male",
      sex %in% c("Other") ~ NA_character_,
      TRUE ~ NA_character_
    ),
    sex_f = factor(sex_std, levels = c("Female", "Male")),
    edu_coll = replace_values(
      caregiver_education,
      c("College", "Some Graduate", "Graduate") ~ "College",
      c("None", "Primary", "Some Secondary", "Secondary", "Some College") ~ "No/some college"
    ) |> factor()
  )

child_o1_clean |>
  summarise(
    n = n(),
    missing_sex = sum(is.na(sex_f)),
    missing_edu = sum(is.na(edu_coll)),
    complete_demo = sum(!is.na(sex_f) & !is.na(edu_ord))
  )
```

```{r}
edu_counts <- child_o1_clean |>
  count(caregiver_education, sort = TRUE)

edu_counts
```

# Multiple imputation (`mice`)

Imputation targets missing `sex_f` and ordered SES (`edu_ord`). Stage-2 outcomes (`theta_est`, SEs, exposure, age) stay observed on analysis rows.

```{r eval=params$run_covariate_models}
demo_unique <- child_o1_clean |>
  distinct(data_id, sex_f, edu_coll)

imp <- mice(
  demo_unique |> select(-data_id),
  m = params$m_imputations,
  maxit = 15,
  seed = 123,
  printFlag = FALSE
)

imp_full <- map(1:params$m_imputations, \(i) {
  comp <- complete(imp, i) |> 
    mutate(data_id = demo_unique$data_id,
           sex_f = factor(sex_f, levels = c("Female", "Male", "Other")),
           edu_coll = factor(edu_coll, levels = c("No/some college", "College")))
  
  child_o1_clean |>
    dplyr::select(-sex_f, -edu_ord, -edu_coll, -sex_std, -sex, -caregiver_education) |>
    left_join(
      comp |> dplyr::select(data_id, sex_f, edu_coll),
      by = "data_id"
    )
})

fits_mi <- brm_multiple(
  bf(theta_est | mi(theta_se) ~ age_c * exp_c + sex_f + edu_coll +
       (1 | child_id) + (age_c * exp_c | dataset_name * language)),
  data = imp_full,
  family = gaussian(),
  prior = c(
    prior(normal(0, 1), class = Intercept),
    prior(normal(0, 1), class = "b"),
    prior(exponential(1), class = sd)
  ),
  save_pars = save_pars(all = TRUE),
  backend = "cmdstanr",
  chains = 2,
  iter = 1000,
  seed = 123,
  cores = 2
)

# fits_mi <- vector("list", params$m_imputations)
# for (i in seq_len(params$m_imputations)) {
#   comp <- complete(imp, i) |>
#     mutate(data_id = demo_unique$data_id) |>
#     mutate(
#       sex_f = factor(sex_f, levels = c("Female", "Male", "Other")),
#       edu_ord = factor(edu_ord, levels = edu_levels, ordered = TRUE)
#     )
# 
#   dat_i <- child_o1_clean |>
#     dplyr::select(-sex_f, -edu_ord, -edu_std, -sex_std, -sex, -caregiver_education) |>
#     left_join(
#       comp |> dplyr::select(data_id, sex_f, edu_ord),
#       by = "data_id"
#     )
# 
#   fits_mi[[i]] <- brm(
#     bf(theta_est | mi(theta_se) ~ age_c * exp_c + sex_f + mo(edu_ord) +
#       (1 | child_id) + (age_c * exp_c | dataset_name * language)),
#     data = dat_i,
#     family = gaussian(),
#     prior = c(
#       prior(normal(0, 1), class = Intercept),
#       prior(normal(0, 1), class = "b"),
#       prior(exponential(1), class = sd),
#       prior(dirichlet(2), class = "simo", coef = "moedu_ord1")
#     ),
#     save_pars = save_pars(all = TRUE),
#     backend = "cmdstanr",
#     chains = 2,
#     iter = 1000,
#     seed = 123,
#     cores = 2
#   )
# }

fit_mi_combined <- combine_models(fits_mi, check_data = F)
saveRDS(fit_mi_combined, here("models", "stage2_o1_covariates_mi.rds"))
```

When `params$run_covariate_models` is `true`, the MI and complete-case chunks evaluate (long runtime).

```{r eval=params$run_covariate_models}
summary(fit_mi_combined)
```

# Complete-case sensitivity

```{r eval=params$run_covariate_models}
child_cc <- child_o1_clean |>
  filter(!is.na(sex_f), !is.na(edu_coll))

fit_cc <- brm(
  bf(theta_est | mi(theta_se) ~ age_c * exp_c * (sex_f + edu_coll) +
    (1 | child_id) + (age_c * exp_c | dataset_name * language)),
  data = child_cc,
  family = gaussian(),
  prior = c(
    prior(normal(0, 1), class = Intercept),
    prior(normal(0, 1), class = "b"),
    prior(exponential(1), class = sd)
  ),
  save_pars = save_pars(all = TRUE),
  backend = "cmdstanr",
  chains = 4,
  iter = 2000,
  cores = 4,
  seed = 124,
  threads = threading(2)
)

saveRDS(fit_cc, here("models", "stage2_o1_covariates_complete_case.rds"))
```

```{r eval=FALSE}
## Inspect saved fits after running chunks above
fit_cc <- readRDS(here("models", "stage2_o1_covariates_complete_case.rds"))
summary(fit_cc)
```

# Comparison to baseline stage-2 model

```{r}
baseline <- readRDS(here(
  "models",
  glue("bilingual_hirt_stage2_model_{params$stage2_variant}.rds")
))

fixef(baseline)
```

Posterior summaries for covariate-augmented models should be compared to `baseline` fixed effects for `age_c`, `exp_c`, and their interaction once MI / complete-case fits are saved.
