---
title: "O2 Lexical class subset models — exploratory"
format: html
params:
  stage2_variant: "linear"
  fit_stage_models: false
---

```{r, include=FALSE}
library(glue)
library(here)
library(brms)
library(cmdstanr)
library(patchwork)
library(tidyverse)

theme_set(
  theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    )
)

knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Item table (`df_data`)

```{r}
bilingual_item_data_clean <- readRDS(here("data", "items.rds"))

f_instruments <- list.files(here("instruments"))
all_instruments <- map(f_instruments, \(f) {
  read_csv(here("instruments", f), show_col_types = FALSE) |>
    mutate(uid = glue("item_{row_number()}")) |>
    pivot_longer(
      cols = -c(category, definition, gloss, uni_lemma, uid),
      names_to = "form",
      values_to = "item_id"
    ) |>
    mutate(
      language = str_remove(f, ".csv"),
      uid = glue("{language}_{uid}")
    )
}) |>
  list_rbind() |>
  filter(!is.na(item_id)) |>
  mutate(form = case_when(
    form == "OxfordExtended" ~ "Oxford Extended",
    form == "OxfordShortII" ~ "Oxford Short II",
    form == "WSExtended" ~ "WS Extended",
    form == "WGShort" ~ "WG Short",
    .default = form
  ))

df_data <- bilingual_item_data_clean |>
  rename(unilemma = uni_lemma) |>
  filter(
    item_kind == "word",
    language != "Other Langs"
  ) |>
  left_join(
    all_instruments |> dplyr::select(language, form, item_id, uid),
    by = join_by(language, form, item_id)
  ) |>
  mutate(
    uid = coalesce(uid, glue("{language}_{item_id}")),
    value = as.numeric(produces),
    age_sc = scale(age)[, 1]
  ) |>
  filter(!is.na(value))
```

# Derived lexical class `lex_class_o2`

Rules: `action_words` → verbs; `descriptive_words` → adjectives; `adverbs` → other; nouns / function_words / verbs / adjectives from `lexical_category` where not superseded; remaining predicates → other; lexical `other` → other.

```{r}
df_items_lex <- df_data |>
  mutate(
    lex_class_o2 = case_when(
      category == "action_words" ~ "verbs",
      category == "descriptive_words" ~ "adjectives",
      lexical_category == "adverbs" ~ "other",
      lexical_category == "nouns" ~ "nouns",
      lexical_category == "function_words" ~ "function_words",
      lexical_category == "verbs" ~ "verbs",
      lexical_category == "adjectives" ~ "adjectives",
      lexical_category == "predicates" ~ "other",
      lexical_category == "other" ~ "other",
      TRUE ~ NA_character_
    )
  )

df_na_check <- df_items_lex |>
  filter(is.na(lex_class_o2)) |>
  dplyr::select(
    language, form, item_id, uid, category,
    lexical_category, lex_class_o2
  )

readr::write_csv(df_na_check, here("data", "lexical_class_na_check.csv"))

count(df_items_lex, lex_class_o2, sort = TRUE)
```

Quality-control file of unresolved rows: [`data/lexical_class_na_check.csv`](data/lexical_class_na_check.csv).

# Stage 1 per lexical class

```{r eval=params$fit_stage_models}
stage1_formula <- bf(
  value ~ exp(logalpha) * (theta + beta),
  theta ~ 0 + (1 | data_id),
  beta ~ 0 + (1 |i| uid),
  logalpha ~ 1 + (1 |i| uid),
  nl = TRUE
)

stage1_priors <- c(
  prior(normal(0, 1), class = "b", nlpar = "logalpha"),
  prior(normal(0, 1), class = "sd", nlpar = "theta"),
  prior(normal(0, 1), class = "sd", nlpar = "beta"),
  prior(normal(0, 1), class = "sd", nlpar = "logalpha")
)

lex_classes <- c("nouns", "verbs", "adjectives", "function_words")

for (lc in lex_classes) {
  df_sub <- df_items_lex |> filter(lex_class_o2 == lc)
  if (nrow(df_sub) < 1000) {
    warning("Very few rows for ", lc)
  }
  mod <- brm(
    formula = stage1_formula,
    data = df_sub,
    family = bernoulli(link = "logit"),
    prior = stage1_priors,
    algorithm = "meanfield",
    iter = 10000,
    tol_rel_obj = 1e-4,
    backend = "cmdstanr",
    threads = threading(2),
    seed = 123
  )
  saveRDS(mod, here("models", glue("stage1_lexical_{lc}.rds")))
}
```

Set chunk `eval: true` locally after reviewing subset sizes; variational fits are slow.

# Stage 2 per lexical class

This follows the same stage-2 setup as `02_exposure_models.qmd`, but uses \(\hat{\theta}\) estimates derived from a stage-1 fit on a lexical-class item subset.

```{r eval=params$fit_stage_models}
df_demogs <- readRDS(here("data", "demographics.rds"))

theta_from_stage1 <- function(mod) {
  theta_ranef <- ranef(mod, summary = TRUE)$data_id[, , "theta_Intercept"]
  theta_fixed <- 0
  tibble(
    data_id = rownames(theta_ranef),
    theta_est = theta_fixed + theta_ranef[, "Estimate"],
    theta_se = theta_ranef[, "Est.Error"]
  )
}

make_child_data_for_theta <- function(theta_estimates, ds_lang_lookup) {
  df_demogs |>
    left_join(ds_lang_lookup, by = "data_id") |>
    mutate(age_sc = scale(age)[, 1]) |>
    dplyr::select(data_id, child_id, age_sc, dataset_name, language, age, languages) |>
    mutate(
      data_id = as.character(data_id),
      child_id = as.character(child_id)
    ) |>
    unnest(languages) |>
    filter(!is.na(exposure_language), !is.na(exposure_proportion)) |>
    dplyr::select(-age_first_exposed) |>
    filter(
      str_detect(language, glue("^{exposure_language}")) |
        exposure_language == "Chinese" & str_detect(language, "^Mandarin")
    ) |>
    mutate(exp_sc = scale(log1p(exposure_proportion / 100))[, 1]) |>
    left_join(theta_estimates, by = "data_id") |>
    filter(exposure_proportion <= 100, !is.na(theta_est))
}

stage2_fit <- function(child_data_cleaned, variant) {
  if (variant == "linear") {
    brm(
      bf(theta_est | mi(theta_se) ~ age_c * exp_c +
        (1 | child_id) + (age_c * exp_c | dataset_name * language)),
      data = child_data_cleaned,
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), class = Intercept),
        prior(normal(0, 1), class = b),
        prior(exponential(1), class = sd)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else if (variant == "quadratic") {
    brm(
      bf(theta_est | mi(theta_se) ~ age_c * poly(exp_c, 2) +
        (1 | child_id) + (age_c * exp_c | dataset_name * language)),
      data = child_data_cleaned,
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), class = Intercept),
        prior(normal(0, 1), class = b),
        prior(exponential(1), class = sd)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else if (variant == "cubic") {
    brm(
      bf(theta_est | mi(theta_se) ~ age_c * poly(exp_c, 3) +
        (1 | child_id) + (age_c * exp_c | dataset_name * language)),
      data = child_data_cleaned,
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), class = Intercept),
        prior(normal(0, 1), class = b),
        prior(exponential(1), class = sd)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else if (variant == "exp") {
    brm(
      bf(theta_est | mi(theta_se) ~ age_c * exp_exp_c +
        (1 | child_id) + (age_c * exp_exp_c | dataset_name * language)),
      data = child_data_cleaned |>
        mutate(exp_exp_c = scale(exp(exposure_proportion), scale = FALSE)),
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), class = Intercept),
        prior(normal(0, 1), class = b),
        prior(exponential(1), class = sd)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else if (variant == "log") {
    brm(
      bf(theta_est | mi(theta_se) ~ age_c * log_exp_c +
        (1 | child_id) + (age_c * log_exp_c | dataset_name * language)),
      data = child_data_cleaned |>
        mutate(log_exp_c = scale(log(exposure_proportion + .01), scale = FALSE)),
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), class = Intercept),
        prior(normal(0, 1), class = b),
        prior(exponential(1), class = sd)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else if (variant == "pw") {
    brm(
      bf(
        theta_est | mi(theta_se) ~ b0 + b1 * age_c +
          b2 * (exp_c - omega) * step(omega - exp_c) +
          b3 * (exp_c - omega) * step(exp_c - omega) +
          b4 * age_c * (exp_c - omega) * step(omega - exp_c) +
          b5 * age_c * (exp_c - omega) * step(exp_c - omega),
        b0 ~ 1 + (1 | child_id),
        b1 + b2 + b3 + b4 + b5 + alpha ~ 1 + (1 | dataset_name * language),
        nlf(omega ~ inv_logit(alpha) - 0.553),
        nl = TRUE
      ),
      data = child_data_cleaned,
      family = gaussian(),
      prior = c(
        prior(normal(0, 1), nlpar = "b0"),
        prior(normal(0, 1), nlpar = "b1"),
        prior(normal(0, 1), nlpar = "b2"),
        prior(normal(0, 1), nlpar = "b3"),
        prior(normal(0, 1), nlpar = "b4"),
        prior(normal(0, 1), nlpar = "b5"),
        prior(normal(0, 1), nlpar = "alpha"),
        prior(exponential(1), class = sd, nlpar = "b0"),
        prior(exponential(1), class = sd, nlpar = "b1"),
        prior(exponential(1), class = sd, nlpar = "b2"),
        prior(exponential(1), class = sd, nlpar = "b3"),
        prior(exponential(1), class = sd, nlpar = "b4"),
        prior(exponential(1), class = sd, nlpar = "b5"),
        prior(exponential(1), class = sd, nlpar = "alpha"),
        prior(exponential(1), class = sigma)
      ),
      save_pars = save_pars(all = TRUE),
      backend = "cmdstanr",
      chains = 4,
      iter = 2000,
      cores = 4,
      seed = 123
    )
  } else {
    stop("Unknown stage2_variant: ", variant)
  }
}

lex_classes <- c("nouns", "verbs", "adjectives", "function_words")

for (lc in lex_classes) {
  message("Lexical class: ", lc)

  stage1_path <- here("models", glue("stage1_lexical_{lc}.rds"))
  if (!file.exists(stage1_path)) {
    warning("Missing stage-1 model for ", lc, " at ", stage1_path)
    next
  }

  mod1 <- readRDS(stage1_path)
  theta_estimates <- theta_from_stage1(mod1)

  ds_lang_lookup <- df_items_lex |>
    filter(lex_class_o2 == lc) |>
    distinct(data_id, dataset_name, language) |>
    mutate(data_id = as.character(data_id))

  child_lc <- make_child_data_for_theta(theta_estimates, ds_lang_lookup)

  child_lc_clean <- child_lc |>
    filter(age >= 7) |>
    mutate(
      exposure_proportion = exposure_proportion / 100,
      exp_c = scale(exposure_proportion, scale = FALSE)[, 1],
      age_c = scale(age, scale = FALSE)[, 1]
    )

  fit2 <- stage2_fit(child_lc_clean, params$stage2_variant)
  saveRDS(
    fit2,
    here("models", glue("stage2_lexical_{lc}_{params$stage2_variant}.rds"))
  )
}
```

# Compare to pooled stage-2 baseline

```{r}
baseline_stage2 <- readRDS(here(
  "models",
  glue("bilingual_hirt_stage2_model_{params$stage2_variant}.rds")
))

fixef(baseline_stage2)
```

When lexical-specific stage-2 fits exist, compare their population-level effects to the pooled baseline.

```{r}
lex_classes <- c("nouns", "verbs", "adjectives", "function_words")

read_lex_fit <- function(lc, variant) {
  p <- here("models", glue("stage2_lexical_{lc}_{variant}.rds"))
  if (!file.exists(p)) return(NULL)
  readRDS(p)
}

lex_fixef <- map(lex_classes, \(lc) {
  mod <- read_lex_fit(lc, params$stage2_variant)
  if (is.null(mod)) return(NULL)
  fx <- fixef(mod) |> as.data.frame()
  fx$term <- rownames(fx)
  tibble(lex_class = lc) |>
    bind_cols(as_tibble(fx))
}) |>
  list_rbind()

if (nrow(lex_fixef) == 0) {
  message("No lexical stage-2 models found to compare.")
} else {
  lex_fixef |>
    filter(term %in% c("Intercept", "age_c", "exp_c", "age_c:exp_c")) |>
    select(lex_class, term, Estimate, Est.Error, Q2.5, Q97.5) |>
    arrange(term, lex_class)
}
```
