---
title: "O5 Dimensionality of bilingual lexical ability — confirmatory mIRT"
format: html
params:
  language_a: "English (American)"
  language_b: "Spanish (Mexican)"
  min_items_per_child: 20
  thin_items: 1          # keep every k-th item per language while prototyping (1 = all)
  fit_models: false
  run_cv: false
  cv_folds: 5
---

# Question

For a bilingual child, consider *every* word they know across both languages. Is
that lexical ability **unidimensional** (one latent trait, languages differing
only in exposure-driven mean level) or **multidimensional** (separable language
lexicons, and/or semantic / lexical-class sub-structure)?

Naively column-binding the two CDIs and running an *exploratory* mIRT will almost
always "find" ≥2 factors, because differential exposure alone forces English
items and Spanish items onto separate factors regardless of the latent structure
of ability. So this file is **confirmatory**: a small set of theoretically
motivated structures, compared with cross-validated held-out log-likelihood and
limited-information fit indices, with exposure modelled as a latent regressor so
the "language factor" is not just soaking up dominance. Read the results against
`10_o5_dimensionality_calibration.qmd`.

::: {.callout-warning}
**On the real EN–SP data this route does not converge.** With 474 bilingual
children and ~1600 items, unregularised 2PL mIRT (even thinned to ~280 items)
runs for hours and hits the EM cycle limit with degenerate loadings. This file
works on the synthetic data in `10` and documents the confirmatory logic, but the
**trustworthy real-data estimate comes from the Bayesian model in
`09_o5_dimensionality_bayes.qmd`**, whose priors regularise the discriminations.
Use `08` for the framing and the calibration pipeline, `09` for the number.
:::

```{r setup, include=FALSE}
library(glue)
library(here)
library(mirt)
library(splines)
library(furrr)
library(tidyverse)

source(here("scripts", "dim_irt_helpers.R"))

theme_set(
  theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    )
)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Data — child-level wide response matrix

Rows are **children** administered in *both* languages of the pair; columns are
`uid`s from both languages; items off the child's form are `NA`. Item difficulty
outliers and thin items are dropped (they manufacture a spurious 2PL factor).

```{r}
langs <- c(params$language_a, params$language_b)

item_table <- build_item_table(
  items       = readRDS(here("data", "items.rds")),
  instruments = load_instruments()
)
df_demogs <- readRDS(here("data", "demographics.rds"))

pair <- build_pair_matrix(
  item_table, df_demogs, langs,
  min_items_per_child = params$min_items_per_child
)

# optional item thinning for fast prototyping of the >2-factor models
if (params$thin_items > 1) {
  keep_uid <- pair$item_lookup |>
    group_by(language) |>
    filter(row_number() %% params$thin_items == 0) |>
    pull(uid)
  pair$resp        <- pair$resp[, keep_uid, drop = FALSE]
  pair$item_lookup <- pair$item_lookup |> filter(uid %in% keep_uid) |>
    arrange(match(uid, colnames(pair$resp)))
}

resp    <- pair$resp
covdata <- as.data.frame(pair$covdata[, c("age_c", "exp_a_c")])
il      <- pair$item_lookup
J       <- ncol(resp)

c(children = nrow(resp), items = J,
  items_a = sum(il$language == params$language_a),
  items_b = sum(il$language == params$language_b))
count(il, language, lex_class)
```

# Model space

| id | structure | "unidimensional" iff |
|----|-----------|----------------------|
| `M0` | unidimensional 2PL, all items | (baseline) |
| `M_lang` | bifactor: general + one specific factor per **language** | ECV → 1, ω_h → ω, implied `r(F_A,F_B)` → 1 |
| `M_class` | bifactor: general + one specific factor per **lexical class** (crossing languages) | class-specific loadings → 0 |

`M_lang` is the primary model. It nests the "one general lexical-ability
dimension" and "two separable lexicons" cases and reports both **ECV / ω_h** and
the **implied cross-language correlation** — so it does the job the
correlated-two-factor model would, without that model's convergence problems
(with per-factor `ns(age)` latent regression it runs ~20 min and does *not*
converge in mirt 1.46.1). `M_class` is the substantively interesting alternative:
if a *predicate* factor spans English and Spanish predicates, the
multidimensionality is **semantic**, not linguistic. Lexical class is collapsed
to 4 levels (nouns / predicates / closed-class / other) — each specific factor is
another EM quadrature dimension, so the full Wordbank scheme is not tractable.

**mirt constraints that shape this (1.46.1):** latent regression (`formula`)
requires an EM-family method, so `method = "EM"`; `bfactor()` does not support
`formula` at all, so the bifactor models are built from explicit `mirt.model()`
specs. `M_lang` (3-D) fits in ~15 min for ~130 items and scales roughly linearly
in item count; `M_class` (4-D) is similar. Prototype with `thin_items > 1`.

The general factor is regressed on `ns(age_c, 3)`; in `M_lang` the language
specifics are additionally regressed on `exp_a_c` so exposure-driven dominance
does not inflate a specific factor.

```{r}
specs <- o5_model_specs(il$language, il$lex_class)
lang_names  <- make.names(unique(il$language))     # S_ factor names in M_lang
class_names <- make.names(sort(unique(il$lex_class)))
```

```{r fit, eval=params$fit_models}
common <- list(itemtype = "2PL", covdata = covdata, method = "EM",
               technical = list(NCYCLES = 3000), verbose = TRUE)

M0 <- do.call(mirt, c(list(resp, specs$M0, formula = ~ ns(age_c, 3)), common))

# general on age; each language specific on exposure
lang_formula <- c(
  list(G = ~ ns(age_c, 3)),
  setNames(rep(list(~ exp_a_c), length(lang_names)), paste0("S_", lang_names))
)
M_lang <- do.call(mirt, c(list(resp, specs$M_lang, formula = lang_formula), common))

M_class <- do.call(mirt, c(list(resp, specs$M_class, formula = ~ ns(age_c, 3)),
                           common))

saveRDS(list(M0 = M0, M_lang = M_lang, M_class = M_class),
        here("models", glue("o5_mirt_{make.names(params$language_a)}.rds")))
```

```{r}
mods <- readRDS(here("models", glue("o5_mirt_{make.names(params$language_a)}.rds")))
list2env(mods, environment())
walk2(names(mods), mods,
      \(nm, m) message(nm, ": converged = ", extract.mirt(m, "converged")))
```

# In-sample comparison

```{r}
anova(M0, M_lang)
anova(M0, M_class)
tibble(
  model  = names(mods),
  npars  = map_dbl(mods, \(m) extract.mirt(m, "nest")),
  AIC    = map_dbl(mods, \(m) extract.mirt(m, "AIC")),
  BIC    = map_dbl(mods, \(m) extract.mirt(m, "BIC")),
  logLik = map_dbl(mods, \(m) extract.mirt(m, "logLik"))
)
```

**Don't stop here** — more factors always fit better in-sample, and wide-range
binary items manufacture a difficulty factor. Use the cross-validated
log-likelihood below, calibrated by file 10.

# Limited-information fit (M2*)

`mirt::M2()` does **not** support latent-regression models (1.46.1), so this only
runs if you refit the models without `formula` (e.g. within a narrow age band).
Left here as a `tryCatch` so it degrades gracefully.

```{r}
m2_fit <- map(mods, \(m) tryCatch(as_tibble(M2(m, type = "C2", calcNull = FALSE)),
                                  error = \(e) tibble(note = conditionMessage(e))))
bind_rows(m2_fit, .id = "model")
```

# Key quantities

```{r}
# --- M_lang: ECV / omega_h (general factor is column 1 of the explicit spec) ---
bifactor_ecv_omega(M_lang, general = 1)

# --- M_lang: correlation between the two language-ability composites ---
bifactor_cross_corr(M_lang, il$language, general = 1)

# --- M_class: is there semantic (lexical-class) sub-structure? ---
bifactor_ecv_omega(M_class, general = 1)
```

Interpretation targets (compare every number to its **null band** in file 10
before concluding — the pipeline mildly *over*estimates ECV):

- **M_lang** `implied_r` > ~.9 and `ECV` > ~.85, `omega_h` close to
  `omega_total` ⇒ one general dimension carries almost all reliable variance;
  languages differ only by exposure-driven level.
- **M_lang** `implied_r` clearly < ~.7 / low `ECV` ⇒ separable lexicons.
- **M_class** class-specific loadings small / `ECV` high ⇒ no meaningful
  *semantic* multidimensionality either.

# Cross-validated held-out log-likelihood

Fits item + latent-regression parameters on training children, **fixes** them,
then evaluates the observed-data log-likelihood on held-out children. Each fold
is a full refit — with `M_lang` at ~15 min that is ~75 min for 5 folds, so run
this with `thin_items` set (e.g. 3–4).

```{r cv, eval=params$run_cv}
plan(multisession, workers = min(params$cv_folds, availableCores()))

specfuns <- list(
  M0 = \(d, cv, ...) mirt(d, specs$M0, itemtype = "2PL", method = "EM", covdata = cv,
                          formula = ~ ns(age_c, 3), verbose = FALSE, ...),
  M_lang = \(d, cv, ...) mirt(d, specs$M_lang, itemtype = "2PL", method = "EM",
                              covdata = cv, formula = lang_formula,
                              verbose = FALSE, ...)
)

cv_res <- imap_dfr(specfuns, \(fn, nm) tibble(
  model = nm,
  cv_loglik = cv_loglik_mirt(resp, covdata, fn, K = params$cv_folds)
))
cv_res |> arrange(desc(cv_loglik))
```

# Exploratory anchor (report, don't lead with)

Age-banded tetrachoric parallel analysis: within a narrow age band the
age-driven common variance is removed, so the retained-factor count speaks to
*residual* structure. Heuristic only.

```{r anchor, eval=params$fit_models}
age_raw <- pair$covdata$age
bands <- tibble(centre = quantile(age_raw, c(.35, .5, .65)) |> round()) |>
  mutate(lo = centre - 3, hi = centre + 3)

pa_by_band <- pmap_dfr(bands, function(centre, lo, hi) {
  rows <- which(age_raw >= lo & age_raw <= hi)
  m <- resp[rows, , drop = FALSE]
  m <- m[, colMeans(!is.na(m)) > .8, drop = FALSE]
  m <- m[, apply(m, 2, \(x) {p <- mean(x, na.rm = TRUE); p > .05 & p < .95}), drop = FALSE]
  pa <- try(psych::fa.parallel(m, cor = "tet", fa = "fa", plot = FALSE,
                               n.iter = 20), silent = TRUE)
  tibble(age_centre = centre, n = length(rows), n_items = ncol(m),
         n_factors = if (inherits(pa, "try-error")) NA_integer_ else pa$nfact)
})
pa_by_band
```

# Caveats

1. **Difficulty-factor artifact** — quantified for this exact pipeline in file 10.
2. **Exposure sparsity** — near-0%-exposure children barely identify that
   language's specific factor.
3. **Age nonlinearity** — `ns(age_c, 3)` in the latent regression is a
   linear-on-θ approximation; refit within 6-month bands as robustness (and to
   get `M2()` limited-info fit indices, which need non-regression models).
4. **`"Log-likelihood was decreasing near the ML solution"`** prints on nearly
   every fit here — it is the known benign mirt EM behaviour with latent
   regression and many parameters, not a failure. Check
   `extract.mirt(m, "converged")` and, for `M_lang`, that the general-factor
   slopes are sane (not all ~6, which would be the degenerate Guttman fit you get
   with no age control).
5. **Pipeline over-estimates ECV** — in synthetic tests `ecv_hat` ≈ 0.81–0.84
   when the truth was 0.65–0.70. Always read the real-data number against its
   file-10 null/calibration band.
6. **Sample size** — EN–SP is likely the only pair with enough bilinguals.
7. Report **ECV / ω_h / implied r** as continuous answers, not a binary verdict.
