---
title: "O5 Dimensionality of bilingual lexical ability — confirmatory mIRT"
format: html
params:
  language_a: "English (American)"
  language_b: "Spanish (Mexican)"
  min_items_per_child: 20
  thin_items: 1          # keep every k-th item per language while prototyping (1 = all)
  fit_models: false
  run_cv: false
  cv_folds: 5
---

# Question

For a bilingual child, consider *every* word they know across both languages. Is
that lexical ability **unidimensional** (one latent trait, languages differing
only in exposure-driven mean level) or **multidimensional** (separable language
lexicons, and/or semantic / lexical-class sub-structure)?

Naively column-binding the two CDIs and running an *exploratory* mIRT will almost
always "find" ≥2 factors, because differential exposure alone forces English
items and Spanish items onto separate factors regardless of the latent structure
of ability. So this file is **confirmatory**: a small set of theoretically
motivated structures, compared with cross-validated held-out log-likelihood and
limited-information fit indices, with exposure modelled as a latent regressor so
the "language factor" is not just soaking up dominance. Read the results against
`10_o5_dimensionality_calibration.qmd`.

```{r setup, include=FALSE}
library(glue)
library(here)
library(mirt)
library(splines)
library(furrr)
library(tidyverse)

source(here("scripts", "dim_irt_helpers.R"))

theme_set(
  theme_bw() +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_blank(),
      strip.background = element_blank(),
      legend.position = "bottom"
    )
)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Data — child-level wide response matrix

Rows are **children** administered in *both* languages of the pair; columns are
`uid`s from both languages; items off the child's form are `NA`. Item difficulty
outliers and thin items are dropped (they manufacture a spurious 2PL factor).

```{r}
langs <- c(params$language_a, params$language_b)

item_table <- build_item_table(
  items       = readRDS(here("data", "items.rds")),
  instruments = load_instruments(here("instruments"))
)
df_demogs <- readRDS(here("data", "demographics.rds"))

pair <- build_pair_matrix(
  item_table, df_demogs, langs,
  min_items_per_child = params$min_items_per_child
)

# optional item thinning for fast prototyping of the >2-factor models
if (params$thin_items > 1) {
  keep_uid <- pair$item_lookup |>
    group_by(language) |>
    filter(row_number() %% params$thin_items == 0) |>
    pull(uid)
  pair$resp        <- pair$resp[, keep_uid, drop = FALSE]
  pair$item_lookup <- pair$item_lookup |> filter(uid %in% keep_uid) |>
    arrange(match(uid, colnames(pair$resp)))
}

resp    <- pair$resp
covdata <- as.data.frame(pair$covdata[, c("age_c", "exp_a_c")])
il      <- pair$item_lookup
J       <- ncol(resp)

c(children = nrow(resp), items = J,
  items_a = sum(il$language == params$language_a),
  items_b = sum(il$language == params$language_b))
count(il, language, lex_class)
```

# Model space

| id | structure | "unidimensional" iff |
|----|-----------|----------------------|
| `M0` | unidimensional 2PL, all items | (baseline) |
| `M1` | two correlated factors by **language** | `cor(F_A, F_B) → 1` |
| `M2` | bifactor: general + **lexical-class** specifics (crossing languages) | class-specific loadings → 0 |
| `M3` | bifactor: general + **language** specifics | ECV → 1, ω_h → ω |

`M1` and `M3` are near-reparameterisations for a 2-language pair (Schmid–Leiman);
`M3` is fit to read off ECV / ω_h. `M2` is the substantively interesting
alternative — if a *noun* factor spans English and Spanish nouns, the
multidimensionality is **semantic**, not linguistic. Every factor is regressed on
`ns(age_c, 3)`; the two language factors additionally on `exp_a_c`.

```{r}
is_a       <- il$language == params$language_a
class_idx  <- as.integer(factor(il$lex_class))    # specific-factor membership (M2)
lang_idx   <- as.integer(factor(il$language))     # specific-factor membership (M3)

spec_M0 <- mirt.model(glue("F = 1-{J}"))
spec_M1 <- mirt.model(glue(
  "FA = {paste(which(is_a),  collapse = ',')}\n",
  "FB = {paste(which(!is_a), collapse = ',')}\n",
  "COV = FA*FB"
))
```

```{r fit, eval=params$fit_models}
common <- list(
  itemtype = "2PL", covdata = covdata,
  technical = list(NCYCLES = 3000, removeEmptyRows = FALSE), verbose = TRUE
)

M0 <- do.call(mirt, c(list(resp, spec_M0, formula = ~ ns(age_c, 3)), common))

M1 <- do.call(mirt, c(list(
  resp, spec_M1, method = "QMCEM",
  formula = list(FA = ~ ns(age_c, 3) + exp_a_c,
                 FB = ~ ns(age_c, 3) + exp_a_c)
), common))

M2 <- do.call(bfactor, c(list(
  resp, model = class_idx, model2 = glue("G = 1-{J}"),
  formula = ~ ns(age_c, 3)
), common))

M3 <- do.call(bfactor, c(list(
  resp, model = lang_idx, model2 = glue("G = 1-{J}"),
  formula = ~ ns(age_c, 3)
), common))

saveRDS(
  list(M0 = M0, M1 = M1, M2 = M2, M3 = M3),
  here("models", glue("o5_mirt_{make.names(params$language_a)}.rds"))
)
```

```{r}
mods <- readRDS(here("models", glue("o5_mirt_{make.names(params$language_a)}.rds")))
list2env(mods, environment())
```

# In-sample comparison

```{r}
anova(M0, M1)          # 1 vs 2 correlated language factors
anova(M0, M3)          # 1 vs bifactor-by-language
anova(M0, M2)          # 1 vs bifactor-by-lexical-class
tibble(
  model  = names(mods),
  npars  = map_dbl(mods, \(m) extract.mirt(m, "nest")),
  AIC    = map_dbl(mods, \(m) extract.mirt(m, "AIC")),
  BIC    = map_dbl(mods, \(m) extract.mirt(m, "BIC")),
  logLik = map_dbl(mods, \(m) extract.mirt(m, "logLik"))
)
```

**Don't stop here** — more factors always fit better in-sample, and wide-range
binary items manufacture a difficulty factor. Use the limited-information indices
and CV below, calibrated by file 10.

# Limited-information fit (M2*)

```{r}
map_dfr(mods, \(m) as_tibble(M2(m, type = "C2", calcNull = FALSE)), .id = "model") |>
  select(model, M2, df, p, RMSEA, CFI, TLI, SRMSR)
```

# Key quantities

```{r}
# --- M1: latent correlation between the two language abilities ---
fcov <- coef(M1, simplify = TRUE)$cov          # estimated factor covariance
cor_ab <- cov2cor(fcov)["FA", "FB"]
cor_ab

# --- M2 / M3: standardised-loading ECV and omega (Stucky & Edelen) ---
bifactor_ecv_omega(M3, general = "G")
bifactor_ecv_omega(M2, general = "G")
```

Interpretation targets:

- `cor_ab` from **M1** > ~.9 ⇒ effectively unidimensional across languages.
- **M3** `ECV` > ~.85 and `omega_h` close to `omega_total` ⇒ one general
  dimension carries almost all reliable variance.
- **M2** class-specific loadings small / `ECV` high ⇒ no meaningful *semantic*
  multidimensionality either.

Compare `cor_ab` / `ECV` to their **null bands** in file 10 before concluding.

# Cross-validated held-out log-likelihood

Fits item + latent-regression parameters on training children, **fixes** them,
then evaluates the observed-data log-likelihood on held-out children — the honest
way to compare dimensionalities.

```{r cv, eval=params$run_cv}
plan(multisession, workers = min(params$cv_folds, availableCores()))

specfuns <- list(
  M0 = \(d, cv, ...) mirt(d, spec_M0, itemtype = "2PL", covdata = cv,
                          formula = ~ ns(age_c, 3), verbose = FALSE, ...),
  M1 = \(d, cv, ...) mirt(d, spec_M1, itemtype = "2PL", covdata = cv, method = "QMCEM",
                          formula = list(FA = ~ ns(age_c, 3) + exp_a_c,
                                         FB = ~ ns(age_c, 3) + exp_a_c),
                          verbose = FALSE, ...),
  M3 = \(d, cv, ...) bfactor(d, lang_idx, model2 = glue("G = 1-{J}"),
                             covdata = cv, formula = ~ ns(age_c, 3),
                             verbose = FALSE, ...)
)

cv_res <- imap_dfr(specfuns, \(fn, nm) tibble(
  model = nm,
  cv_loglik = cv_loglik_mirt(resp, covdata, fn, K = params$cv_folds)
))
cv_res |> arrange(desc(cv_loglik))
```

# Exploratory anchor (report, don't lead with)

Age-banded tetrachoric parallel analysis: within a narrow age band the
age-driven common variance is removed, so the retained-factor count speaks to
*residual* structure. Heuristic only.

```{r anchor, eval=params$fit_models}
age_raw <- pair$covdata$age
bands <- tibble(centre = quantile(age_raw, c(.35, .5, .65)) |> round()) |>
  mutate(lo = centre - 3, hi = centre + 3)

pa_by_band <- pmap_dfr(bands, function(centre, lo, hi) {
  rows <- which(age_raw >= lo & age_raw <= hi)
  m <- resp[rows, , drop = FALSE]
  m <- m[, colMeans(!is.na(m)) > .8, drop = FALSE]
  m <- m[, apply(m, 2, \(x) {p <- mean(x, na.rm = TRUE); p > .05 & p < .95}), drop = FALSE]
  pa <- try(psych::fa.parallel(m, cor = "tet", fa = "fa", plot = FALSE,
                               n.iter = 20), silent = TRUE)
  tibble(age_centre = centre, n = length(rows), n_items = ncol(m),
         n_factors = if (inherits(pa, "try-error")) NA_integer_ else pa$nfact)
})
pa_by_band
```

# Caveats

1. **Difficulty-factor artifact** — quantified for this exact pipeline in file 10.
2. **Exposure sparsity** — near-0%-exposure children barely identify that
   language's specific factor.
3. **Age nonlinearity** — `ns(age_c, 3)` is a linear-on-θ approximation; refit
   within 6-month bands as robustness.
4. **Sample size** — EN–SP is likely the only pair with enough bilinguals.
5. Report **ECV / ω_h / cor(F_A,F_B)** as continuous answers, not a binary verdict.
