---
title: "O5 Dimensionality — Bayesian crossed random-effects IRT"
format: html
params:
  language_a: "English (American)"
  language_b: "Spanish (Mexican)"
  min_items_per_child: 20
  thin_items: 1           # keep every k-th item/language; full EN-SP is ~1665 items
  fit_models: false
  method: "sample"        # "sample" (NUTS). ADVI/pathfinder are unreliable here — see below.
  link_concepts: "auto"   # "auto" | TRUE | FALSE — TE linking only helps with lots of overlap
---

::: {.callout-important}
**Use NUTS, not ADVI.** On the real EN–SP data (474 children) rstan's meanfield
VB diverged completely — ELBO oscillating, `rho`/`ecv`/`lambda_g` collapsing to 0.
NUTS on the same model is well-behaved (0 divergences, coherent posterior). The
mirt route in `08` also fails to converge on this sample. This file is the one
that produces a trustworthy number, at the cost of a multi-hour run.
:::

# Idea

Rather than let differential exposure masquerade as a factor (as an exploratory
mIRT would), model it, and turn the dimensionality question into a
**variance-components** question:

- item easiness = translation-equivalent **concept** term (shared by `dog` /
  `perro` via `uni_lemma`) + **language-specific offset**;
- child ability `theta_child[i, L] = lambda_g * theta_g[i] + theta_s[i, L]`,
  `theta_g ~ N(0,1)`, `theta_s[,L] ~ N(0, sigma_s[L])`;
- exposure + age enter as fixed per-language effects.

Then

- `sigma_s -> 0`  ⇒ unidimensional  (`rho -> 1`)
- `lambda_g -> 0` ⇒ two independent lexicons (`rho -> 0`)
- both free       ⇒ estimate `rho` and `ECV` directly — a **continuous** answer.

`models/bilingual_dim_irt.stan` implements all three via the data flags
`estimate_general` / `estimate_specific` (it is a crossed random-effects
generalisation of the old concept/word IRT in `archive/models/model1.stan`).

```{r setup, include=FALSE}
library(glue)
library(here)
library(cmdstanr)
library(posterior)
library(loo)
library(tidyverse)

source(here("scripts", "dim_irt_helpers.R"))

theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_blank(),
          legend.position = "bottom")
)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Data prep (shares helpers with `08_o5_dimensionality_mirt.qmd`)

```{r}
langs <- c(params$language_a, params$language_b)

item_table <- build_item_table(
  items       = readRDS(here("data", "items.rds")),
  instruments = load_instruments()
)
df_demogs <- readRDS(here("data", "demographics.rds"))

pair <- build_pair_matrix(
  item_table, df_demogs, langs,
  min_items_per_child = params$min_items_per_child
)

if (params$thin_items > 1) {
  keep_uid <- pair$item_lookup |> group_by(language) |>
    filter(row_number() %% params$thin_items == 0) |> pull(uid)
  pair$resp        <- pair$resp[, keep_uid, drop = FALSE]
  pair$item_lookup <- pair$item_lookup |> filter(uid %in% keep_uid) |>
    arrange(match(uid, colnames(pair$resp)))
}

# long form the Stan model wants: one row per (child, item)
obs <- wide_to_long_obs(pair$resp, pair$item_lookup, pair$covdata)
exposure <- child_language_exposure(df_demogs, langs)

link <- if (identical(params$link_concepts, "auto")) NA else
          as.integer(isTRUE(params$link_concepts))

prep <- build_dim_irt_stan_data(
  obs, exposure, langs,
  estimate_general = 1L, estimate_specific = 1L,
  link_concepts = link, compute_loglik = 1L
)
str(prep$data[c("N", "I", "J", "C", "link_concepts")])
```

# Fit the three nested variants

Runtime scales with item count. Rough NUTS cost (4 chains, 1000+1000): ~1 h at
150 items, ~3–4 h at 400, 8 h+ at the full 1665. Prototype with
`thin_items: 4`; do the headline run unthinned overnight. The variance
components (`sigma_delta`, `sigma_s`) mix slowest — watch their `rhat` / `ess`.

```{r fit, eval=params$fit_models}
mod <- cmdstan_model(here("models", "bilingual_dim_irt.stan"))

run_variant <- function(g, s, tag) {
  d <- modifyList(prep$data, list(estimate_general = g, estimate_specific = s))
  fit <- mod$sample(data = d, seed = 123, chains = 4, parallel_chains = 4,
                    iter_warmup = 1000, iter_sampling = 1000,
                    adapt_delta = 0.95, max_treedepth = 12)
  fit$save_object(here("models", glue("o5_bayes_{tag}_{make.names(params$language_a)}.rds")))
  print(fit$diagnostic_summary())
  fit
}

fit_full <- run_variant(1, 1, "full")      # general + specific  -> rho, ecv
fit_uni  <- run_variant(1, 0, "unidim")    # sigma_s == 0        -> nested null
fit_two  <- run_variant(0, 1, "twolex")    # lambda_g == 0       -> nested null
```

::: {.callout-note}
**Result, EN (American) × ES (Mexican), 474 children, 554 items** (thinned 1-in-3
from 1665; `link_concepts = 0`; 4 chains × 1000, 0 divergences, `rho`/`ecv`
`n_eff` ≈ 1300, `rhat` = 1.00):

| quantity | median | 95% CI |
|---|---|---|
| `rho`  (cross-language ability correlation) | **0.585** | 0.519–0.642 |
| `ecv`  (person-level explained common variance) | **0.582** | 0.515–0.640 |
| `lambda_g` | 1.85 | 1.67–2.03 |
| `sigma_s` (EN / ES specific SD) | 1.74 / 1.37 | — |

A **strong common lexical-ability dimension (~58% of reliable person variance)
plus substantial, genuinely language-specific ability (~40%)**. The posterior
excludes both `rho`→1 (single dimension) and `rho`→0 (independent lexicons); the
`unidim` and `twolex` nested fits are not credible. `beta_exp` ≈ 5.4 / 3.4
(exposure strongly shifts which language's words a child produces); `beta_age`
≈ 0.15 [−0.05, 0.35] (weak once exposure + general ability are in — a spline on
age would be a cleaner robustness check).

Still to do: `10`'s calibration (SBC on this model — the pipeline may *over*state
`rho`); a full-item run; a `link_concepts = 1` robustness fit on the ~95
translation-equivalent items; other language pairs.

Saved: `models/o5_bayes_EN-SP_summary.csv`, `models/o5_bayes_EN-SP_full_draws.csv`.
:::

```{r}
rd <- \(tag) readRDS(here("models", glue("o5_bayes_{tag}_{make.names(params$language_a)}.rds")))
fit_full <- rd("full"); fit_uni <- rd("unidim"); fit_two <- rd("twolex")
```

# The answer, read straight off the full model

```{r}
fit_full$summary(c("lambda_g", "sigma_s", "rho", "ecv",
                   "beta_exp", "beta_age", "sigma_concept", "sigma_delta"))

as_draws_df(fit_full$draws(c("rho", "ecv"))) |>
  pivot_longer(c(rho, ecv)) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 60) +
  facet_wrap(~name, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "Posterior: correlation of the two language abilities, and person-level ECV")
```

- `rho` near 1 and `ecv` near 1 ⇒ **unidimensional**: one general trait,
  languages differ only by exposure-driven level.
- `rho` clearly below ~.7, `sigma_s` away from 0 ⇒ **separable lexicons**.
- Intermediate `ecv` (~.6–.85) ⇒ strong common trait **plus** non-trivial
  language-specific ability — report the number, don't binarise.

# Model comparison (LOO), as a cross-check

```{r}
loo_compare(list(
  full   = fit_full$loo(),
  unidim = fit_uni$loo(),
  twolex = fit_two$loo()
))
```

If `unidim` is within ~2 SE of `full`, the language-specific factors aren't
earning their keep. If `full` clearly beats both, general + specific is the right
description.

# Robustness

- **Age.** Re-fit within 6-month age bands (drop `beta_age*`), confirm
  `rho` / `ecv` are stable.
- **Concept linking.** Compare to `link_concepts: false` — if `rho` drops a lot,
  translation-equivalent structure is doing real work (conceptual vs lexical
  representation).
- **NUTS vs Pathfinder.** Iterate with `method: "pathfinder"` but report the NUTS
  fit — VI understates `sigma_s` uncertainty, the parameter in question.
- Feed `theta_child[,1]` / `theta_child[,2]` into `02_exposure_models.qmd` as a
  consistency check against the two-stage pipeline.

# SBC calibration (answers `10`'s open question)

One-rep-per-condition recovery check: simulate CDI responses from a **known**
`rho_true`, using the real EN–SP form design / age / exposure and empirical item
difficulties (from marginal endorsement rates) + discriminations drawn to match
the fitted `sigma_logalpha` (~0.27), then refit the full model.

```{r, eval=FALSE}
# models/o5_bayes_SBC_calibration.csv
#  rho_true  rho_hat_median   95% CI            covers?
#      0.30      0.279       [0.191, 0.364]        yes
#      0.60      0.576       [0.512, 0.636]        yes
#      0.90      0.882       [0.858, 0.902]        yes  (Rhat 1.01, weaker ESS)
```

Recovery is good across the range, with a small (~0.02) low-side bias and no
systematic inflation. **This means the headline `rho`/`ecv` numbers below can be
read at face value** — the earlier worry (carried over from the mirt route in
`08`, which *does* overstate multidimensionality-adjacent quantities) does not
transfer to this Stan model.

# Cross-pair replication

| pair | children | items | `rho` (median, 95% CI) | `ecv` | `beta_age` |
|---|---|---|---|---|---|
| English (American) × Spanish (Mexican) | 474 | 554 | 0.585 [0.519, 0.642] | 0.582 | 0.15 [−0.05, 0.35] |
| English (British) × Spanish (European) | 690 | ~420 | 0.708 [0.664, 0.747] | 0.706 | 2.21 [2.03, 2.39] |

Both pairs land in the same qualitative regime — decisively **rejecting both**
`rho → 1` (unidimensional) **and** `rho → 0` (separable lexicons) — but the
British/European pair shows a *higher* shared-variance share (~71% vs ~58%) and
a much stronger residual age effect. Whether that's a real difference between
these two bilingual populations (different countries, dataset mixes, age
ranges) or an artifact of the item thinning / sample composition is open; the
Malaysian English/Malay/Mandarin trilingual sample (n=569, would need `L=3` in
the Stan model) and a full-item, unthinned refit of both pairs are the natural
next checks.

Saved: `models/o5_bayes_ENGB-ESEU_summary.csv`, `models/o5_bayes_SBC_calibration.csv`.
