---
title: "O5 Dimensionality — Bayesian crossed random-effects IRT"
format: html
params:
  language_a: "English (American)"
  language_b: "Spanish (Mexican)"
  min_items_per_child: 20
  fit_models: false
  method: "sample"        # "sample" (NUTS) or "pathfinder"
  link_concepts: true     # FALSE -> every item its own concept (robustness check)
---

# Idea

Rather than let differential exposure masquerade as a factor (as an exploratory
mIRT would), model it, and turn the dimensionality question into a
**variance-components** question:

- item easiness = translation-equivalent **concept** term (shared by `dog` /
  `perro` via `uni_lemma`) + **language-specific offset**;
- child ability `theta_child[i, L] = lambda_g * theta_g[i] + theta_s[i, L]`,
  `theta_g ~ N(0,1)`, `theta_s[,L] ~ N(0, sigma_s[L])`;
- exposure + age enter as fixed per-language effects.

Then

- `sigma_s -> 0`  ⇒ unidimensional  (`rho -> 1`)
- `lambda_g -> 0` ⇒ two independent lexicons (`rho -> 0`)
- both free       ⇒ estimate `rho` and `ECV` directly — a **continuous** answer.

`models/bilingual_dim_irt.stan` implements all three via the data flags
`estimate_general` / `estimate_specific` (it is a crossed random-effects
generalisation of the old concept/word IRT in `archive/models/model1.stan`).

```{r setup, include=FALSE}
library(glue)
library(here)
library(cmdstanr)
library(posterior)
library(loo)
library(tidyverse)

source(here("scripts", "dim_irt_helpers.R"))

theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_blank(),
          legend.position = "bottom")
)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Data prep (shares helpers with `08_o5_dimensionality_mirt.qmd`)

```{r}
langs <- c(params$language_a, params$language_b)

item_table <- build_item_table(
  items       = readRDS(here("data", "items.rds")),
  instruments = load_instruments()
)
df_demogs <- readRDS(here("data", "demographics.rds"))

pair <- build_pair_matrix(
  item_table, df_demogs, langs,
  min_items_per_child = params$min_items_per_child
)

# long form the Stan model wants: one row per (child, item)
obs <- wide_to_long_obs(pair$resp, pair$item_lookup, pair$covdata)
if (!params$link_concepts) obs <- obs |> mutate(unilemma = NA_character_)

exposure <- child_language_exposure(df_demogs, langs)

prep <- build_dim_irt_stan_data(
  obs, exposure, langs,
  estimate_general = 1L, estimate_specific = 1L, compute_loglik = 1L
)
str(prep$data[c("N", "I", "J", "C")])
```

# Fit the three nested variants

```{r fit, eval=params$fit_models}
mod <- cmdstan_model(here("models", "bilingual_dim_irt.stan"))

run_variant <- function(g, s, tag) {
  d <- modifyList(prep$data, list(estimate_general = g, estimate_specific = s))
  fit <- if (params$method == "pathfinder") {
    mod$pathfinder(data = d, seed = 123, draws = 2000, num_paths = 8)
  } else {
    mod$sample(data = d, seed = 123, chains = 4, parallel_chains = 4,
               iter_warmup = 750, iter_sampling = 750, adapt_delta = 0.95)
  }
  fit$save_object(here("models", glue("o5_bayes_{tag}_{make.names(params$language_a)}.rds")))
  fit
}

fit_full <- run_variant(1, 1, "full")      # general + specific
fit_uni  <- run_variant(1, 0, "unidim")    # sigma_s == 0
fit_two  <- run_variant(0, 1, "twolex")    # lambda_g == 0
```

```{r}
rd <- \(tag) readRDS(here("models", glue("o5_bayes_{tag}_{make.names(params$language_a)}.rds")))
fit_full <- rd("full"); fit_uni <- rd("unidim"); fit_two <- rd("twolex")
```

# The answer, read straight off the full model

```{r}
fit_full$summary(c("lambda_g", "sigma_s", "rho", "ecv",
                   "beta_exp", "beta_age", "sigma_concept", "sigma_delta"))

as_draws_df(fit_full$draws(c("rho", "ecv"))) |>
  pivot_longer(c(rho, ecv)) |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 60) +
  facet_wrap(~name, scales = "free") +
  labs(x = NULL, y = NULL,
       title = "Posterior: correlation of the two language abilities, and person-level ECV")
```

- `rho` near 1 and `ecv` near 1 ⇒ **unidimensional**: one general trait,
  languages differ only by exposure-driven level.
- `rho` clearly below ~.7, `sigma_s` away from 0 ⇒ **separable lexicons**.
- Intermediate `ecv` (~.6–.85) ⇒ strong common trait **plus** non-trivial
  language-specific ability — report the number, don't binarise.

# Model comparison (LOO), as a cross-check

```{r}
loo_compare(list(
  full   = fit_full$loo(),
  unidim = fit_uni$loo(),
  twolex = fit_two$loo()
))
```

If `unidim` is within ~2 SE of `full`, the language-specific factors aren't
earning their keep. If `full` clearly beats both, general + specific is the right
description.

# Robustness

- **Age.** Re-fit within 6-month age bands (drop `beta_age*`), confirm
  `rho` / `ecv` are stable.
- **Concept linking.** Compare to `link_concepts: false` — if `rho` drops a lot,
  translation-equivalent structure is doing real work (conceptual vs lexical
  representation).
- **NUTS vs Pathfinder.** Iterate with `method: "pathfinder"` but report the NUTS
  fit — VI understates `sigma_s` uncertainty, the parameter in question.
- Feed `theta_child[,1]` / `theta_child[,2]` into `02_exposure_models.qmd` as a
  consistency check against the two-stage pipeline.
