---
title: "O5 Dimensionality — simulation calibration"
format: html
params:
  language_a: "English (American)"
  language_b: "Spanish (Mexican)"
  min_items_per_child: 20
  n_rep: 100
  dev_sd: 0.6            # SD of the structure-controlled ability deviation
  run_sim: false
  run_bayes_sbc: false
---

# Why this file exists

`08_o5_dimensionality_mirt.qmd` and `09_o5_dimensionality_bayes.qmd` each return
*some* estimate of `cor(F_A, F_B)`, `ECV`, and a preferred model. Before believing
any of it:

1. **False-positive multidimensionality.** If the truth is a single latent trait,
   how often does the pipeline nonetheless prefer ≥2 factors — from
   exposure-driven sparsity, the CDI age range, and the binary-item
   difficulty-factor artifact? What do `cor(F_A,F_B)` / `ECV` look like when the
   true value is 1?
2. **Power & bias.** If the truth *is* two partly-separable lexicons (`rho < 1`)
   or a bifactor structure (`ECV < 1`), can we detect it, and is the estimate
   biased?

We simulate responses from **known** structures, reusing the **real** item
parameters (`data/item_est.csv`), the **real** child age/exposure values, and the
**real** form design (which items each child saw), then run the exact pipeline
from file 08.

```{r setup, include=FALSE}
library(glue)
library(here)
library(mirt)
library(splines)
library(furrr)
library(tidyverse)

source(here("scripts", "dim_irt_helpers.R"))

theme_set(
  theme_bw() +
    theme(panel.grid.minor = element_blank(),
          panel.grid.major = element_blank(),
          strip.background = element_blank(),
          legend.position = "bottom")
)
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

# Ingredients

```{r}
langs <- c(params$language_a, params$language_b)

item_table <- build_item_table(
  items       = readRDS(here("data", "items.rds")),
  instruments = load_instruments(here("instruments"))
)
df_demogs  <- readRDS(here("data", "demographics.rds"))
item_est   <- read_csv(here("data", "item_est.csv"), show_col_types = FALSE)
child_data <- readRDS(here("data", "bilingual_hirt_stage2_data.rds"))

# real design: children in both languages, real form (NA pattern), real covariates
pair <- build_pair_matrix(item_table, df_demogs, langs,
                          min_items_per_child = params$min_items_per_child)

# item params aligned to the columns of pair$resp; drop items with no estimate
ipar <- tibble(uid = colnames(pair$resp)) |>
  left_join(item_est |> select(uid, a = logalpha_est, b = beta_est), by = "uid") |>
  mutate(a = exp(coalesce(a, 0)), b = coalesce(b, mean(b, na.rm = TRUE)))

# real per-language ability level per child (sets the age-driven mean + scale)
mu_tbl <- child_data |>
  filter(language %in% langs) |>
  transmute(child_id = as.character(child_id), language, theta_real = theta_est) |>
  pivot_wider(names_from = language, values_from = theta_real) |>
  rename(mu_a = !!langs[1], mu_b = !!langs[2])

design <- tibble(child_id = rownames(pair$resp)) |>
  left_join(mu_tbl, by = "child_id") |>
  left_join(pair$covdata |> select(child_id, exp_a_c), by = "child_id") |>
  mutate(mu_a = coalesce(mu_a, mean(mu_a, na.rm = TRUE)),
         mu_b = coalesce(mu_b, mean(mu_b, na.rm = TRUE)))

is_a <- pair$item_lookup$language == langs[1]
col_lang <- pair$item_lookup$language
```

# Generative model

`produces ~ Bernoulli(inv_logit(a_j * (theta_child[i, L(j)] + b_j)))` on the cells
the real child actually saw. `theta_child[i, L] = mu_i,L + dev[i, L]`, where
`mu_i,L` is the child's real per-language ability level (carries age + real
exposure) and `dev` is the mean-zero, SD-`dev_sd` piece whose structure we
control:

| `structure` | `dev` | truth |
|-------------|-------|-------|
| `unidim`     | `dev[,1] = dev[,2] = z_g` | `rho = 1`, `ECV = 1` |
| `twolexicon` | `MVN(0, [[1, rho], [rho, 1]]) * dev_sd` | given `rho` |
| `bifactor`   | `sqrt(ECV)*z_g + sqrt(1-ECV)*z_s`, scaled | given `ECV` |

```{r}
simulate_rep <- function(structure, rho = 1, ecv = 1, seed = 1) {
  set.seed(seed)
  n <- nrow(pair$resp); s <- params$dev_sd

  dev <- switch(structure,
    unidim = { g <- rnorm(n); cbind(g, g) * s },
    twolexicon = {
      MASS::mvrnorm(n, c(0, 0), matrix(c(1, rho, rho, 1), 2)) * s
    },
    bifactor = {
      g <- rnorm(n); z <- cbind(rnorm(n), rnorm(n))
      (sqrt(ecv) * cbind(g, g) + sqrt(1 - ecv) * z) * s
    },
    stop("unknown structure")
  )

  abil_a <- design$mu_a + dev[, 1]
  abil_b <- design$mu_b + dev[, 2]
  abil   <- ifelse(matrix(is_a, nrow = n, ncol = length(is_a), byrow = TRUE),
                   abil_a, abil_b)              # n x J, ability for each cell

  eta <- sweep(abil, 2, ipar$b, `+`)
  eta <- sweep(eta, 2, ipar$a, `*`)
  P   <- plogis(eta)
  y   <- matrix(rbinom(length(P), 1, P), nrow = n, dimnames = dimnames(pair$resp))
  y[is.na(pair$resp)] <- NA                     # keep the real form design
  y
}
```

# Pipeline under test (mirror of file 08: M0 vs M1)

```{r}
analyse_rep <- function(y, truth) {
  # drop items that came out near-constant in this replicate
  ok <- apply(y, 2, \(x) { p <- mean(x, na.rm = TRUE)
                           sum(!is.na(x)) >= 50 && p > .02 && p < .98 })
  y  <- y[, ok, drop = FALSE]
  isa <- (col_lang[ok] == params$language_a)
  J   <- ncol(y)
  cov <- as.data.frame(pair$covdata[, c("age_c", "exp_a_c")])

  spec_M0 <- mirt.model(glue("F = 1-{J}"))
  spec_M1 <- mirt.model(glue(
    "FA = {paste(which(isa),  collapse = ',')}\n",
    "FB = {paste(which(!isa), collapse = ',')}\nCOV = FA*FB"))

  M0 <- mirt(y, spec_M0, itemtype = "2PL", covdata = cov,
             formula = ~ ns(age_c, 3), verbose = FALSE)
  M1 <- mirt(y, spec_M1, itemtype = "2PL", method = "QMCEM", covdata = cov,
             formula = list(FA = ~ ns(age_c, 3) + exp_a_c,
                            FB = ~ ns(age_c, 3) + exp_a_c), verbose = FALSE)

  cor_ab <- cov2cor(coef(M1, simplify = TRUE)$cov)["FA", "FB"]
  lr <- anova(M0, M1)

  tibble(
    structure  = truth$structure,
    rho_true   = truth$rho,
    ecv_true   = truth$ecv,
    cor_ab_hat = cor_ab,
    lr_p       = lr$p[2],
    prefer_2f  = lr$p[2] < 0.05 &&
                 extract.mirt(M1, "BIC") < extract.mirt(M0, "BIC")
  )
}
```

# Run the grid

```{r sim, eval=params$run_sim}
plan(multisession, workers = max(1, availableCores() - 1))

grid <- bind_rows(
  tibble(structure = "unidim",     rho = 1,             ecv = 1),
  tibble(structure = "twolexicon", rho = c(.3, .6, .9), ecv = NA_real_),
  tibble(structure = "bifactor",   rho = NA_real_,      ecv = c(.5, .7, .9))
) |>
  crossing(rep = seq_len(params$n_rep)) |>
  mutate(seed = row_number() + 1000)

results <- future_pmap_dfr(grid, function(structure, rho, ecv, rep, seed) {
  y <- simulate_rep(structure, rho = coalesce(rho, 1),
                    ecv = coalesce(ecv, 1), seed = seed)
  analyse_rep(y, list(structure = structure, rho = rho, ecv = ecv))
}, .options = furrr_options(seed = TRUE))

saveRDS(results, here("models", glue("o5_calibration_{make.names(params$language_a)}.rds")))
```

```{r}
results <- readRDS(here("models", glue("o5_calibration_{make.names(params$language_a)}.rds")))
```

# 1. False-positive multidimensionality (truth = unidimensional)

```{r}
results |>
  filter(structure == "unidim") |>
  summarise(
    p_prefer_2factor   = mean(prefer_2f),
    cor_ab_hat_median  = median(cor_ab_hat),        # should be ~1
    cor_ab_hat_q05     = quantile(cor_ab_hat, .05),
    cor_ab_hat_q95     = quantile(cor_ab_hat, .95)
  )
```

This is the **null calibration** of the whole pipeline. If `p_prefer_2factor` is,
say, 0.30, a "2-factor wins" result on real data is not evidence of
multidimensionality on its own, and the real-data `cor_ab_hat` must be compared
against the `[q05, q95]` null band — not against 1.0.

# 2. Power and bias (truth = separable / bifactor)

```{r}
results |>
  filter(structure != "unidim") |>
  group_by(structure, rho_true, ecv_true) |>
  summarise(
    power_detect   = mean(prefer_2f),
    cor_ab_hat_mean = mean(cor_ab_hat),
    cor_ab_hat_bias = mean(cor_ab_hat - rho_true),   # NA for bifactor rows
    .groups = "drop"
  )
```

```{r}
results |>
  filter(structure == "twolexicon") |>
  ggplot(aes(x = factor(rho_true), y = cor_ab_hat)) +
  geom_hline(aes(yintercept = rho_true), lty = "dashed", col = "grey50") +
  geom_violin(fill = "grey85") +
  geom_boxplot(width = .15, outlier.shape = NA) +
  labs(x = "True correlation between language lexicons",
       y = expression(widehat(cor)(F[A], F[B])),
       title = "Calibration of the estimated language-ability correlation")
```

# 3. (Optional) SBC-lite for the Bayesian model

Fit `models/bilingual_dim_irt.stan` (full variant) to a few simulated data sets
per structure; check the `rho` / `sigma_s` posteriors are calibrated — under
`unidim` truth `sigma_s` should pile up near 0 and `rho`'s 90% interval should
cover 1; under `twolexicon` truth `rho`'s interval should cover the truth at
~nominal rate.

```{r bayes_sbc, eval=params$run_bayes_sbc}
library(cmdstanr)
mod <- cmdstan_model(here("models", "bilingual_dim_irt.stan"))
exposure <- child_language_exposure(df_demogs, langs)

sbc_one <- function(structure, rho = 1, ecv = 1, seed = 1) {
  y   <- simulate_rep(structure, rho = rho, ecv = ecv, seed = seed)
  obs <- wide_to_long_obs(y, pair$item_lookup, pair$covdata)
  prep <- build_dim_irt_stan_data(obs, exposure, langs,
                                  estimate_general = 1L, estimate_specific = 1L,
                                  compute_loglik = 0L)
  fit <- mod$sample(data = prep$data, chains = 2, parallel_chains = 2,
                    iter_warmup = 500, iter_sampling = 500, seed = seed,
                    refresh = 0)
  d <- posterior::as_draws_df(fit$draws(c("rho", "ecv", "sigma_s[1]", "sigma_s[2]")))
  tibble(structure, rho_true = rho, ecv_true = ecv,
         rho_med   = median(d$rho),
         rho_cover = quantile(d$rho, .05) <= rho & rho <= quantile(d$rho, .95),
         sigma_s1_med = median(d$`sigma_s[1]`))
}

sbc_res <- bind_rows(
  map_dfr(1:8,  \(i) sbc_one("unidim", seed = i)),
  map_dfr(1:8,  \(i) sbc_one("twolexicon", rho = .6, seed = 100 + i))
)
sbc_res |>
  group_by(structure) |>
  summarise(rho_med = mean(rho_med), rho_coverage = mean(rho_cover),
            sigma_s1_med = mean(sigma_s1_med), .groups = "drop")
```

# What to conclude

- Report the real-data `cor(F_A,F_B)` / `ECV` / preferred-model result **next to**
  its null calibration (section 1) and its power (section 2).
- Real-data estimate inside the unidimensional-null band ⇒ "consistent with a
  single general lexical-ability dimension, language differences attributable to
  exposure".
- Outside the null band **and** section 2 shows power at that effect size ⇒ a
  defensible multidimensionality claim, quantified by `ECV`, not asserted.
